Time delay estimation in unresolved lensed quasars

### 2020-12-22 paper/aa-package/paper.tex
intro

@@

\section{Introduction}
The Hubble parameter $H_0$ quantifies the current expansion rate of the Universe. The measured value of $H_0$ from different observations have led to a tension. In particular, EU observations [\cite{planck}] have measured $H_0 = 67.4 \pm 0.5$ km s$^{-1}$ Mpc$^{-1}$, whereas, LU observations [\cite{cepheid_H0}] give $H_0 = 74.03 \pm 1.42$ km s$^{-1}$ Mpc$^{-1}$, resulting in a tension of about $4.4\sigma$.
\\ 
As first pointed out in [\cite{Refsdal1964}], time-delay cosmography offers an alternative way of determining the Hubble parameter: the light rays coming from a distant source, e.g. a quasar, can be deflected by the gravitational field of an intervening massive object, e.g. a galaxy. If the field is strong enough, multiple images of the same source will be observed and, by tracking the light intensity of each image over time, a \emph{light curve} is obtained. Light curves associated with different images will exhibit a mutual time-delay ($\Delta T$), due to the different paths that the photons have travelled. As shown in [\cite{Refsdal1964}], this time-delay is related to the Hubble parameter as $H_0 \propto 1/\Delta T$. The major results obtained via time-delay cosmography come from the H0LiCOW collaboration [\cite{H0licow_XIII}], which has found $H_0 = 73.3_{-1.8}^{+1.7}$ km s$^{-1}$ Mpc$^{-1}$ from a sample of six lensed quasars monitored by the COSMOGRAIL  project [\cite{Cosmograil2020}], enhancing the tension up to $5.3 \sigma$. However, a more recent analysis of 40 strong gravitational lenses, from TDCOSMO+SLACS \cite{tdcosmo}, has found $H_0 = 67.4_{-3.2}^{+4.1}$ km s$^{-1}$ Mpc$^{-1}$, relaxing the Hubble tension and demonstrating the importance of understanding the lenses mass density profiles. This scenario motivates further studies aimed at improving the precision in the $H_0$ estimation.
\\
Time-delays between light curves of Gravitationally Lensed Quasars (GLQs) can be estimated with various methods \cite{Tewes2013}, such as free-knot spline interpolation \cite{free} or via Gaussian Process (GP) regression \cite{Rasmussen}. Such methods work under the assumption that multiple images of the same quasar are fully resolved so that a light curve can be extracted for each of them. However, most known GLQs \cite{quasars_db} have small angular separations $\Delta\theta < \sim{3}{arcsec}$. 
This would make big telescopes the ideal instruments to perform lensed quasars monitoring, both in light of their high angular resolution and the geographical areas they are placed in, where the effects of atmospheric turbulence are less prominent.
%This would make big telescopes, with high angular resolution, the ideal instruments to perform lensed quasars monitoring. Furthermore, these telescopes are placed in geographical areas where effects of atmospheric turbulence are less prominent. 
However, because of the time scales of the intrinsic variations of the source, such observation campaigns should last years \cite{Cosmograil2020}. Therefore, due to the amount of observational requests that big telescopes have to fulfill, they can not be employed for these purposes. On the other hand, small/medium sized telescopes ($1$-$2$m) \cite{Borgeest1996} can be used. Unfortunately their reduced angular resolution, together with their often less privileged geographical positions in terms of atmospheric seeing, may worsen the effective angular resolution up to $3 arcsec$ \cite{karttunen2016fundamental}. In addition, small telescopes cannot observe faint distant sources. These shortcomings prevent the acquisition of a large number of GLQ datasets suitable for the application of the aforementioned techniques \cite{free, Rasmussen}.
Here, we propose a novel approach based on Machine Learning (ML) algorithms to estimate the time-delay from non-resolved lensed quasar light curves. In particular, we train a one-dimensional Convolutional Neural Network (CNN) to map non-resolved double-lensed quasar images into the corresponding time-delay. Starting from open-source real data \cite{CosmograilData} of resolved quasar light curves, we combine them into a single time series to mimic the real situation in which the quasar images are not resolved. Then, we feed the so-obtained time series into a GP-based algorithm, which artificially generates new realistic non-resolved quasar light curves along with their corresponding time-delays. Preliminary experiments based on data of quasars HE0435 and HS2209 demonstrate the effectiveness of the proposed method.

>>>>

 \usepackage{braket}

@@

\section{Introduction}
The Hubble parameter $H_0$ quantifies the current expansion rate of the Universe. The measured value of $H_0$ from different observations have led to a tension. In particular, EU observations [\cite{planck}] have measured $H_0 = 67.4 \pm 0.5$ km s$^{-1}$ Mpc$^{-1}$, whereas, LU observations [\cite{cepheid_H0}] give $H_0 = 74.03 \pm 1.42$ km s$^{-1}$ Mpc$^{-1}$, resulting in a tension of about $4.4\sigma$.
\\ 
As first pointed out in [\cite{Refsdal1964}], time-delay cosmography offers an alternative way of determining the Hubble parameter: the light rays coming from a distant source, e.g. a quasar, can be deflected by the gravitational field of an intervening massive object, e.g. a galaxy. If the field is strong enough, multiple images of the same source will be observed and, by tracking the light intensity of each image over time, a \emph{light curve} is obtained. Light curves associated with different images will exhibit a mutual time-delay ($\Delta T$), due to the different paths that the photons have travelled. As shown in [\cite{Refsdal1964}], this time-delay is related to the Hubble parameter as $H_0 \propto 1/\Delta T$. The major results obtained via time-delay cosmography come from the H0LiCOW collaboration [\cite{H0licow_XIII}], which has found $H_0 = 73.3_{-1.8}^{+1.7}$ km s$^{-1}$ Mpc$^{-1}$ from a sample of six lensed quasars monitored by the COSMOGRAIL  project [\cite{Cosmograil2020}], enhancing the tension up to $5.3 \sigma$. However, a more recent analysis of 40 strong gravitational lenses, from TDCOSMO+SLACS \cite{tdcosmo}, has found $H_0 = 67.4_{-3.2}^{+4.1}$ km s$^{-1}$ Mpc$^{-1}$, relaxing the Hubble tension and demonstrating the importance of understanding the lenses mass density profiles. This scenario motivates further studies aimed at improving the precision in the $H_0$ estimation. 
\\
In this respect, the fractional error of $H_0$, for an ensemble of N Gravitationally Lensed Quasars (GLQs), is related to the uncertainties in the time-delay estimation $\sigma_{\Delta T}$, line-of-sight convergence $\sigma_{los}$ and lens surface density $\sigma_{\braket{k}}$ as 
\begin{equation}
\frac{\sigma_H^2}{H_0^2} \sim \frac{\sigma_{\Delta T}^2 + \sigma_{los}^2 + \sigma_{\braket{k}}^2}{N}
\end{equation}
Therefore, increasing the sample of analysed GLQs allows to reduce the error on $H_0$.

Time-delays between light curves of Gravitationally Lensed Quasars (GLQs) can be estimated with various methods \cite{Tewes2013}, such as free-knot spline interpolation \cite{free} or via Gaussian Process (GP) regression \cite{Rasmussen}. Such methods work under the assumption that multiple images of the same quasar are fully resolved so that a light curve can be extracted for each of them. However, most known GLQs \cite{quasars_db} have small angular separations $\Delta\theta < \sim{3}{arcsec}$. 
This would make big telescopes the ideal instruments to perform lensed quasars monitoring, both in light of their high angular resolution and the geographical areas they are placed in, where the effects of atmospheric turbulence are less prominent.
%This would make big telescopes, with high angular resolution, the ideal instruments to perform lensed quasars monitoring. Furthermore, these telescopes are placed in geographical areas where effects of atmospheric turbulence are less prominent. 
However, because of the time scales of the intrinsic variations of the source, such observation campaigns should last years \cite{Cosmograil2020}. Therefore, due to the amount of observational requests that big telescopes have to fulfill, they can not be employed for these purposes. On the other hand, small/medium sized telescopes ($1$-$2$m) \cite{Borgeest1996} can be used. Unfortunately their reduced angular resolution, together with their often less privileged geographical positions in terms of atmospheric seeing, may worsen the effective angular resolution up to $3 arcsec$ \cite{karttunen2016fundamental}. In addition, small telescopes cannot observe faint distant sources. These shortcomings prevent the acquisition of a large number of GLQ datasets suitable for the application of the aforementioned techniques \cite{free, Rasmussen}.
Here, we propose a novel approach based on Machine Learning (ML) algorithms to estimate the time-delay from non-resolved lensed quasar light curves. In particular, we train a one-dimensional Convolutional Neural Network (CNN) to map non-resolved double-lensed quasar images into the corresponding time-delay. Starting from open-source real data \cite{CosmograilData} of resolved quasar light curves, we combine them into a single time series to mimic the real situation in which the quasar images are not resolved. Then, we feed the so-obtained time series into a GP-based algorithm, which artificially generates new realistic non-resolved quasar light curves along with their corresponding time-delays. Preliminary experiments based on data of quasars HE0435 and HS2209 demonstrate the effectiveness of the proposed method.
